Structural Breaks in Inflation Dynamics within the European Monetary Union

### 2011-05-19 pkg/paper/Paper/draft.Rnw
updated draft
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/glogis@95 b679b2f7-6539-45d0-a605-5ddcb976cde9

## additional functions
cross_country_moments <- function(countries, date,
  probs = c(0.25, 0.5, 0.75))
{
  rval <- sapply(countries, function(country) {
    nb <- results[[country]]$nbreaks
    m <- t(sapply(results[[country]]$model, moments))
    bd <- if(nb < 1) {
      end(results[[country]]$data)
    } else {
      breakdates(results[[country]]$breakpoints, breaks = nb)
    }
    m[sum(as.yearmon(date) > bd) + 1, ]
  })
  apply(rval, 1, quantile, probs = probs)
}

@@

## For the creation of the tables
## <<tabelle, eval=FALSE>>=
root_dir <- do.call("file.path", as.list(head(strsplit(getwd(), .Platform$file.sep)[[1]], -1)))
if(!file.exists(file.path(root_dir, "Analysis", "Austria.rda"))) {
setwd(file.path(root_dir, "Analysis"))
## run results4 (with dummy creation for sales periods)
Sweave("results4.Rnw")
setwd(file.path(root_dir, "Paper"))
}

@@

format_cross_country <- function(crossmom, digits = 6, name="Euro") {
  rval <- format(round(crossmom, digits = digits), nsmall = digits)
  rval <- cbind(name,"", matrix(paste("$", rval, "$", sep = ""), ncol = 3))
  rval[1,1] <- name
  rval[2,1] <- ""
  rval[3,1] <- ""
  rval[1,2] <- "lower"
  rval[2,2] <- "medium"
  rval[3,2] <- "upper"
  rval <- apply(rval, 1, paste, collapse = " & ")
  rval <- paste(rval, "\\\\")
  rval[length(rval)] <- paste(rval[length(rval)], "\\hline")
  return(rval)
}

@@

%(... will jesus früher anetua ...) \\

@@

%(... am Schluss so enden, dass nun interessant, wie sich die Momente so genau verändert haben; auf Basis der theorie--> suche nach brüchen, schiefe noch hinein/besser hinein; 3 saha simultan aschoa...)\\

@@

%Considering the prior knowledge available about the timing of a countries participation in any of the aforementioned steps towards monetary union, an exact timing of breakpoints in the moments of the respective inflation series enables us to draw conclusions about the exact effects of the EMU the inflation rates of its members. 

@@

%(... jesus: maybe comment on results here ...) \\

@@

\clearpage
\section{Data}\label{s2}
%(... first method, than data...) \\

% wichtig: http://sdw.ecb.europa.eu/quickview.do?SERIES_KEY=122.ICP.M.U2.N.000000.4.ANR
% sieht man die daten von sales prices: 1999-1 portugal,  2000-1 lux, 2001-1 bel, 2002 ita-spa (allerdings in anderem paper ein jahr früher) hm

We \fixme{Jesus: yearly inflation rates better; don't start with data in paper} employ seasonally unadjusted HICP series of 21 countries ranging from January 1990 to December 2010 that are obtained from the \citet{oecd}. Countries included are Austria,	Belgium, Czech Republic,	Denmark, Estonia,	Finland,	France,	Germany,	Greece,	Hungary,	Ireland,	Italy,	Luxembourg,	Netherlands,	Poland,	Portugal,	Slovakia, Slovenia,  Spain,	Sweden, and	United Kingdom. Latvia and Lithuania as well as Bulgaria and Romania are excluded due to data scarcity. Cypria and Malta are not included since they are very small economies. \\
The countries in this sample can be divided into three different groups. First the euro countries (Austria, Belgium, Estonia, France, Finland, Germany, Greece, Ireland, Italy, Luxembourg, Netherlands, Portugal, Slovakia, Slovenia and Spain).  Second, the EU members not participating in the ERM~II: Czech Republic, Hungary, Poland, United Kingdom, and Sweden. Denmark stands on its own as a member of the EU and the ERM~II, but not yet a member of the EMU. \\
%(... schlussverkäufe schon hier erwähnen, einfach cool, ja das gibts, wir haben die aber weggeschätzt; generell viel angelsächsischer/präpotenter schreiben; nur 1 weg der richtige und zwar unserer; nicht erwähnen, dass andere dinge auch noch möglich; so nach motto: mich interessierts, ich machs, andere erst am ende erwähnen...)
% greece: -for winter discounts: First 10 days of January-end of February; -for summer discounts: First 10 days of July -end od August; but has always been included
As \readme{Jesus: Italy and Spain are problematic, fixed with same method as Belgium and Luxembourg} shown in a paper by \cite{pltm}, the harmonization of the treatment of sales prices led to considerable changes in some countries (Belgium, Italy, Luxembourg and Spain); failing to take this into account may lead to considerable distortions in the estimation. In these cases, we dealt with the inclusion of the sales periods effects\footnote{This was demanded by a Commission regulation, details may be found in: \citet{eurlex} and in the Appendix \ref{dataprobs}.} into the HICP series (January 2000 for Belgium, January 2001 for Italy, January 1999 for Luxembourg and January 2001 for Spain) by estimating dummies for these month\footnote{There are two month with a significant decrease in prices (January and July), followed by significant price increases in the following months (February and August).} and subtracting them from the original time series.
\fixme{update for using of X-12-ARIMA method, argue with seasonal effects in \cite{pltm}: In Germany and Portugal, sales prices are included from 1998 onwards, in Ireland and Luxembourg from 1999 onwards and in Belgium from 2000. In January 2001, Italy and Spain were the last two EU15 countries to include sales into their HICP statistics (The official date may be 12 months later. However, the indices of the previous 12 months had been revised in order to avoid distortions in the inflation rates.)}

@@

To monitor changes in the dynamics of the inflation rates, we follow the usual strategy employed when testing for structural breaks. First, the model is fitted for the whole sample using the generalized logistic distribution. This then enables testing for the stability of the parameter estimates. If there is evidence of significant changes, the number and the timing of the break points are estimated. \\
This framework allows to trace the evolution of mean, variance and skewness of the inflation series and thus interpreting deviations in any of these parameters as possibly due to a change in monetary policy, is made feasible. Any huge intervention or change is captured by fluctuations in the parameter stability of the series and can be quantified and related to a distinct time frame. \\
%(...discuss correlation...) \\
The correlation \fixme{Jesus: yes, but a DGP needs to be assumed, it would need to take care of any autororrelation, wouldn't it?} dynamics of the series is of less importance to this framework, which concentrates on changes in mean, variance and skewness and does not focus on autocorrelation, the use of alternative tests, as described in \cite{andreou}, is also hindered by the low number of observations, due to the monthly frequencies.\footnote{In some years the Billion Prices Project of the \cite{bpp}, which provides a much higher data frequency, could solve this problem.} 
%To provide consistent estimates of the parameter vector $\hat{\phi}$, the covariance matrix $\hat{V}^{-1/2}$ is estimated using a heteroskedasticity and autocorrelation consistent covariance estimator.

% korrelation besser herausarbeiten
% mehr: link to, match with known events, like intervention

@@

\begin{eqnarray}
f(y | \theta, \sigma, \delta) & = & \frac{\frac{\delta}{\sigma} \cdot \exp^{-\frac{y-\theta}{\sigma}}}{(1+\exp^{-\frac{y-\theta}{\sigma}})^{(\delta+1)}} 
\label{eqn1}
\end{eqnarray}
% skewness: if delta = 1.1 --> skweness is 0.1, is positive, skewed to the right: linkssteil
% ie. negative skew (delta <1: rechtssteil, in englisch: skewed to the left 
% (psigamma(0.5, deriv = 2) - psigamma(1, deriv = 2))/(psigamma(1, deriv = 1) + psigamma(0.5, deriv = 1))^(3/2)

@@

<<echo=FALSE>>=
        x <- results$Slovenia
        slovpval <- sctest(x$efp, functional = supLM(0.1))$p.value
@

To illustrate our method, we consider the case of Slovenia. After a period of relative instability following the independence from Yugoslavia, Slovenia was successfull in realigning its economy and introduced a number of reforms leading to a stable growth in the past years. The good economic performance made it possible for Slovenia to enter the ERM~II in June 2004 and later on to introduce the euro in January 2007. The data collected contains the period from January 1996 onwards. In the first few years, Slovenia experienced very high inflation rates, but was successfull in reducing inflation to a much lower level. \\
A \fixme{Jesus: confidence interval for break} glance at the time series depicted in Figure \ref{fig1} of the monthly inflation rates reveals a decline in the mean around the middle of the observed time period. The $efp(\cdot)$ starts to cross the border as early as 2001 and reaches the peak somewhere between 2003 and 2004. The p--value of the supLM test is \Sexpr{round(slovpval, 5)} and clearly indicates at least one break.  The number of breaks is calculated by means of the LWZ criteria, which chooses one break in August 2003.\footnote{For the changes in the parameters, see Table \ref{resulttable}.} In this year, Slovenia introduced a number of financial reforms and severely decreased the year on year growth in the M3 monetary aggregate. Combined, these actions were enough to ensure the successfull participation in the ERM~II a year later and the introduction of the euro two years later. \readme{Jesus: include information on how the parameters change: I did give a footnote here}

\begin{figure}[h!tp]
<<fig=TRUE, height=4, width=8>>=
x <- results$Slovenia
par(mfrow = c(1, 2))
plot(x$data, main = "Slovenia", ylab = "100 * diff(log(HICP))", xlab = "Time")
lines(x$breakpoints, breaks = x$nbreaks)
lines(fitted(x$breakpoints, breaks = x$nbreaks, type = "mean"), col = 4)
plot(x$efp, functional = supLM(0.1), main = "supLM test")
lines(x$breakpoints, breaks = x$nbreakss)
@
\caption{Slovenia: breakpoint estimate and supLM test}
\label{fig1}
\end{figure}

\newpage
\section{Results}\label{s5}
%(... jesus: show test statistic: no room...) \\
% das package colortbl (\usepackege[dvips]{colortbl}) ermöglicht z.B. mit
%\rowcolor[gray]{0.90} in einer Tabellenzeile diese farbig (bzw. hier grau) zu unterlegen.

\footnotesize
\begin{longtable}{|l|l|r|r|r|l|l|l|}
\hline
\multicolumn{8}{|c|}{Austria and Belgium} \\
\hline
Country & Segment & Mean & Variance & Skewness & ERM & ERM~II & Euro \\
\hline
<<echo=FALSE, results=tex>>=
writeLines(unlist(lapply(c(1, 2), format_moments)))
@
\hline
\multicolumn{8}{|c|}{The Eastern countries} \\
\hline
<<echo=FALSE, results=tex>>=
writeLines(unlist(lapply(c(3,5,10,15,17,18), format_moments)))
@
\hline
\multicolumn{8}{|c|}{Italy and Spain} \\
\hline
<<echo=FALSE, results=tex>>=
writeLines(unlist(lapply(c(12, 19), format_moments)))
@
\hline
\multicolumn{8}{|c|}{Ireland} \\
\hline
<<echo=FALSE, results=tex>>=
writeLines(unlist(lapply(c(11), format_moments)))
@
\hline
\multicolumn{8}{|c|}{No change countries} \\
\hline
<<echo=FALSE, results=tex>>=
writeLines(unlist(lapply(c(9,14,7,6), format_moments)))
@
\hline
\multicolumn{8}{|c|}{The 90s} \\
\hline
<<echo=FALSE, results=tex>>=
writeLines(unlist(lapply(c(16,20, 21), format_moments)))
@
\hline
\multicolumn{8}{|c|}{Former DM--zone} \\
\hline
<<echo=FALSE, results=tex>>=
writeLines(unlist(lapply(c(4,8,13), format_moments)))
@
\caption{Break point, parameter estimates and ERM information}
\label{resulttable}
\end{longtable}
% warning: the groups then change; Italy and Greece have not yet answered
% Italian Statistics: Your request has been correctly inserted in the system with the reference number 13078/2010.
% comment: the greek break is NOT sure, since they have not yet answered about sales periods

@@

%(...3 punkte darstellung noch einbauen...) \\
%(...andere struktur: 1.short introduction, 2. countries in detail (6.1), 3. summary (6.2) ...) \\
%(... jesus: opening confusing and obscure...)\\
\readme{Jesus: include p-values in table; no place left; Jesus: opening seems a bit confusing and obscure} Table \ref{resulttable} presents the break dates and the parameter values of the segments as well as the dates of entry into the ERM/ERM~II and the date of the euro introduction. We formed seven groups of countries according to the observed pattern of break points. These seven groups consist of Austria and Belgium, the Eastern European countries, the two big southern European countries, Italy and Spain, Ireland, the countries with no change in the inflation dynamics, the countries with breaks in the early 90s and the member countries of the former DM--zone (excluding Austria and Belgium). \\

@@

Regarding Austria and Belgium, the time after the break is characterized by an increase in the mean and a severe increase in the variance of the inflation rates. The skewness is now negative for both series, which indicates that the distribution has shifted to the right, implying that overall inflation is likely to increase. This change can be traced back to two major influences: the financial crisis and potentially some spillover effects of higher oil prices in late 2007. \\
The second group consists of the Eastern European countries: Czech Republic, Estonia, Hungary, Poland, Slovakia and Slovenia. In almost all of these countries, the mean and the variance of their inflation rates declined in the later part of the 90s (Slovenia is an exception with an increase in variance after the break). Most of them experienced a break in 1997/1998, with Poland, Slovakia and Slovenia somewhat later. The break during the late 90s is a result of these countries efforts to curb in inflation by decreasing money supply and indicates the time when these countrie's actually overcame the biggest transitionally shocks on their way towards free market economies. Slovakia is very similar to Slovenia, as its inflation dynamics too changed roughly one year prior to its entry into ERM~II. \\
%(... jesus: maybe some reference on MP (what is MP???)...) \\
%(... thinks this is just data artefact, however even after desaisonalization the same effect...) \\
\fixme{Jesus: this is just a data construct: but after desaisonalization still almost the same} Another combination of countries with rather identical results are the two biggest southern European countries, Italy and Spain. Both of them experienced high rates of inflation in the early 90s. Afterwards, in the phase leading up to the monetary union, mean and variance were greatly reduced. Soon after the fixing of the exchange rates, this trend changed significantly, leading to a higher mean and a much higher variance. A possible reason for this is that the fiscal responsibility that was required for an entry into the eurozone was softened after this goal was achieved. \\
%(... in irland der schock schon persistenter als in anderen ländern, bei der grauen graphik zeigen (greyplot...)\\
Ireland can be considered a rather special case. 
%It is the only economy where the effects of the ongoing financial crisis of 2008 are visible.  
In Ireland, compared to other countries, the deflationary effect of the financial crises was more persistent. This deflationary trend, which for a time affected all countries, can be seen in Figure \ref{greyplot}, which depicts a rolling mean estimation, where the mean is calculated using the first moment of a fitted GL distribution. There is evidence of a decline in mean inflation, which returns to its previous path roughly a year later. The break in 2008 can be attributed to the strong contraction triggered by the bursting of the Irish housing bubble and a beginning deflation in concert with the massive costs of the bailout of the countries banking system. Surprisingly, this deflationary pressure is not visible in Spain, which was affected by a housing bubble of roughly the same size as Ireland, however in Spain housing prices did not adjust so abruptly and in such a degree.  \\
%(...change formulation for greece...) \\
Another interesting group consists of four economically rather different countries. In none of them do we find any evidence of a change in inflation dynamics. Whereas France, Finland and Netherlands are on the lower side of the inflation margin, the Greek inflation rate is among the highest.  \\
The next group consists of two countries that decided not to be part of any European Monetary System\footnote{If we ignore the short participating time of both countries in the EMS.}, the United Kingdom and Sweden. Portugal is included due to the same break point behavior. All three show a clear break in the early 90s. Two of them can be traced back to economic crisis, the currency crisis of the United Kingdom cumulating into the ``Black Wednesday'' in 1992 and the run on the Swedish currency in the same year. In the case of Portugal, the inflation rate after 1992 was much more stable, showing that the inflationary problems of Portugal were less severe thereafter.\\
The last group consists of the former DM--zone countries Denmark, Germany and Luxembourg.  In the case of Denmark we observe an increase in variance after 2000 with a very distinct change in skewness. In Germany, the inflation rate increased both on average and in its variability. Luxembourg suffered a severe increase in variance, similar to Italy and Spain. All of these countries, which profited from the stability of the previous German currency, still have a very low inflation rate.

@@

<<echo=FALSE>>=
data("HICP", package = "glogis")
euhi <- HICP[, colnames(HICP) == "EU"]
euhi <- 100 * diff(log(na.omit(euhi)))
euhim1 <- mean(window(euhi, start=yearmon(1990), end= yearmon(1998+11/12)))
euhim2 <- mean(window(euhi, start=yearmon(1998), end=yearmon(2008+7/12)))
euhim3 <- mean(window(euhi, start=yearmon(2008+8/10), end=yearmon(2010+11/12)))
euhiv1 <- var(window(euhi, start=yearmon(1990), end= yearmon(1998+11/12)))
euhiv2 <- var(window(euhi, start=yearmon(1998), end=yearmon(2008+7/12)))
euhiv3 <- var(window(euhi, start=yearmon(2008+8/10), end=yearmon(2010+11/12)))
euhis1 <- skewness(window(euhi, start=yearmon(1990), end= yearmon(1998+11/12)))
euhis2 <- skewness(window(euhi, start=yearmon(1998), end=yearmon(2008+7/12)))
euhis3 <- skewness(window(euhi, start=yearmon(2008+8/10), end=yearmon(2010+11/12)))
@

@@

% if(!file.exists(file.path(root_dir, "Paper", "draft-greyplot.pdf")))

%\begin{figure}[h!tp]
%<<greyplot, fig=TRUE, height=8, width=8>>=
%
% ## compute GL mean (if no NA)
%  glmean <- function(x) {
%    if(any(is.na(x))) NA else glogisfit(x, hessian = FALSE)$moments[1]
%  }
%  hicps <- hicps[, colnames(hicps) != "EU"]
%
%  ## rolling GL means
%  hicp_glmean <- rollapply(hicps, 12, glmean)
%
%  ## semi-transparent gray and red
%  gray_red <- rgb(c(0.2, 0.8), c(0.2, 0), c(0.2, 0), alpha = 0.3)
%
%  ## plot returns (gray) and rolling means (red)
%  plot(hicps, plot.type = "single", lwd = 1.5, col = gray_red[1], xlab="Time", ylab="Series")
%  for(i in 1:ncol(hicp)) lines(hicp_glmean[,i], lwd = 1.5, col = gray_red[2])
%@
%\caption{Inflation rates and estimated GL mean over time}
%\label{greyplot}
%\end{figure}

@@

Regarding the general trend, Table \ref{grouptable} presents lower, medium and upper (25, 50, 75) percentiles of the first three moments of the inflation series  at the beginning of the convergence phase (January 1995), in December 2007 as a point in time before the financial crisis  and for the last observation in the sample. The countries are grouped into the ``core'' of the eurozone (Austria, Belgium, Finland, France, Germany, Luxembourg and Netherlands), the PIIGS countries (Portugal, Italy, Ireland, Greece and Spain) and all the others.  \\

>>>>

## additional functions
cross_country_moments <- function(countries, date=yearmon(c(1990,1995,2000,2010)),
  probs = c(0.25, 0.5, 0.75))
{
ret <- mat.or.vec(9,length(date)) 
for(jj in 1:length(date)) {
  rval <- sapply(countries, function(country) {
    nb <- results[[country]]$nbreaks 
    m <- t(sapply(results[[country]]$model, moments)) 
    bd <- if(nb < 1) { 
      end(results[[country]]$data) 
    } else {
      breakdates(results[[country]]$breakpoints, breaks = nb) 
    }
    m[sum(as.yearmon(date[jj]) > bd) + 1, ] 
  })
  v1 <- apply(rval, 1, quantile, probs = probs)
  ret[1:3,jj] <- v1[,1] 
  ret[4:6,jj] <- v1[,2] 
  ret[7:9,jj] <- v1[,3] 
}
  return(ret)
}

@@

## For the creation of the tables
## <<tabelle, eval=FALSE>>=
root_dir <- do.call("file.path", as.list(head(strsplit(getwd(), .Platform$file.sep)[[1]], -1)))
if(!file.exists(file.path(root_dir, "Analysis", "Austria.rda"))) {
setwd(file.path(root_dir, "Analysis"))
## run results4 (with dummy creation for sales periods)
Sweave("results1s.Rnw")
setwd(file.path(root_dir, "Paper"))
}

@@

format_cross_country <- function(crossmom, digits = 3, moment="mean", quantil="lower") {
  rval <- format(round(crossmom, digits = digits), nsmall = digits)
  rval <- cbind(moment, quantil, matrix(paste("$", rval, "$", sep=""), ncol = 4))
  rval[1,1] <- "mean"
  rval[2:3,1] <- ""
  rval[1,2] <- "lower"
  rval[2,2] <- "medium"
  rval[3,2] <- "upper"
  rval[4,1] <- "variance"
  rval[5:6,1] <- ""
  rval[4,2] <- "lower"
  rval[5,2] <- "medium"
  rval[6,2] <- "upper"
  rval[7,1] <- "skewness"
  rval[8:9,1] <- ""
  rval[7,2] <- "lower"
  rval[8,2] <- "medium"
  rval[9,2] <- "upper"
  rval <- apply(rval, 1, paste, collapse = " & ")
  rval <- paste(rval, "\\\\")
  rval[length(rval)] <- paste(rval[length(rval)], "\\hline")
  return(rval)
}

@@

@@

@@

@@

@@

\clearpage
\section{Data}\label{s2}

We \fixme{Jesus: yearly inflation rates better; don't start with data in paper} employ seasonally unadjusted HICP series of 21 countries ranging from January 1990 to December 2010 that are obtained from the \citet{oecd}. Countries included are Austria,	Belgium, Czech Republic,	Denmark, Estonia,	Finland,	France,	Germany,	Greece,	Hungary,	Ireland,	Italy,	Luxembourg,	Netherlands,	Poland,	Portugal,	Slovakia, Slovenia,  Spain,	Sweden, and	United Kingdom. Latvia and Lithuania as well as Bulgaria and Romania are excluded due to data scarcity. Cypria and Malta are not included since they are very small economies. \\
The countries in this sample can be divided into three different groups. First the euro countries (Austria, Belgium, Estonia, France, Finland, Germany, Greece, Ireland, Italy, Luxembourg, Netherlands, Portugal, Slovakia, Slovenia and Spain).  Second, the EU members not participating in the ERM~II: Czech Republic, Hungary, Poland, United Kingdom, and Sweden. Denmark stands on its own as a member of the EU and the ERM~II, but not yet a member of the EMU. \\
As  shown in a paper by \cite{pltm}, the harmonization of the treatment of sales prices led to considerable changes in some countries (Germany, Portugal, Ireland, Luxembourg, Belgium, Italy and Spain); failing to take this into account may lead to considerable distortions in the estimation. We dealt with the inclusion of sales periods\footnote{This was demanded by a Commission regulation, details may be found in: \citet{eurlex} and in the Appendix \ref{dataprobs}.} and all the other seasonality in the data by using the X--12--ARIMA method for desaisonalization, as described in \cite{X-12-ARIMA} and \cite{Findley+Monsell+Bell:1998}.

@@

To monitor changes in the dynamics of the inflation rates, we follow the usual strategy employed when testing for structural breaks. First, the model is fitted for the whole sample using the generalized logistic distribution. This then enables testing for the stability of the parameter estimates. If there is evidence of significant changes, the number and the timing of the break points are estimated. \\
This framework allows to trace the evolution of mean, variance and skewness of the inflation series and thus interpreting deviations in any of these parameters as possibly due to a change in monetary policy, is made feasible. Any huge intervention or change is captured by fluctuations in the parameter stability of the series and can be quantified and related to a distinct time frame. \\
The correlation \fixme{Jesus: yes, but a DGP needs to be assumed, it would need to take care of any autororrelation, wouldn't it?} dynamics of the series is of less importance to this framework, which concentrates on changes in mean, variance and skewness and does not focus on autocorrelation, the use of alternative tests, as described in \cite{andreou}, is also hindered by the low number of observations, due to the monthly frequencies.\footnote{In some years the Billion Prices Project of the \cite{bpp}, which provides a much higher data frequency, could solve this problem.}

@@

\begin{eqnarray}
f(y | \theta, \sigma, \delta) & = & \frac{\frac{\delta}{\sigma} \cdot \exp^{-\frac{y-\theta}{\sigma}}}{(1+\exp^{-\frac{y-\theta}{\sigma}})^{(\delta+1)}} 
\label{eqn1}
\end{eqnarray}

@@

<<echo=FALSE>>=
        x <- results$Slovenia
        spval <- sctest(x$efp, functional = supLM(0.1))$p.value
@

To illustrate our method, we consider the case of Slovenia. After a period of relative instability following the independence from Yugoslavia, Slovenia was successfull in realigning its economy and introduced a number of reforms leading to a stable growth in the past years. The good economic performance made it possible for Slovenia to enter the ERM~II in June 2004 and later on to introduce the euro in January 2007. The data collected contains the period from January 1996 onwards. In the first few years, Slovenia experienced very high inflation rates, but was successfull in reducing inflation to a much lower level. \\
A glance at the time series depicted in Figure \ref{fig1} of the monthly inflation rates reveals a decline in the mean around the middle of the observed time period. The $efp(\cdot)$ starts to cross the border as early as 2001 and reaches the peak somewhere between 2003 and 2004. The p--value of the supLM test is \Sexpr{round(spval, 8)} and clearly indicates at least one break.  The number of breaks is calculated by means of the LWZ criteria, which chooses one break in August 2003. The goodness of fit test depicted in Figure \ref{figtest} confirms the choice of the GL distribution. The mean inflation rate was cut to only a third of its previous value, whereas variance increased slightly and the distribution slightly skewed to the left, but peaking at a lower value than previously. \\
In this year, Slovenia introduced a number of financial reforms and severely decreased the year on year growth in the M3 monetary aggregate. Combined, these actions were enough to ensure the successfull participation in the ERM~II a year later and the introduction of the euro two years later.

\begin{figure}[h!tp]
<<fig=TRUE, height=4, width=8>>=
x <- results$Slovenia
par(mfrow = c(1, 2))
plot(x$data, main = "Slovenia", ylab = "100 * diff(log(HICP))", xlab = "Time")
lines(x$breakpoints, breaks = x$nbreaks)
lines(fitted(x$breakpoints, breaks = x$nbreaks, type = "mean"), col = 4)
if(x$nbreaks > 0) lines(confint(x$breakpoints, breaks = x$nbreaks))
plot(x$efp, functional = supLM(0.1), main = "supLM test")
lines(x$breakpoints, breaks = x$nbreakss)
@
\caption{Slovenia: breakpoint estimate and supLM test}
\label{fig1}
\end{figure}

\begin{figure}[h!tp]
<<fig=TRUE, height=4, width=8>>=
x <- results$Slovenia
par(mfrow=c(1,2))
for(i in 1:2)
plot(x$model[[i]], moments = TRUE,
  xlim = c(-1.5, 2), ylim = c(0, 1.3),
  breaks = seq(-1.5, 2, by = 0.25),
  xlab = paste(names(x$model)[i], "\n", "Goodness-of-fit p-value: ",
    format.pval(summary(x$model[[i]])$chisq.test$p.value, digits = 4),
    sep = ""))
@
\caption{Slovenia: Goodness of fit test}
\label{figtest}
\end{figure}

\clearpage
\newpage
\section{Results}\label{s5}

\footnotesize
\begin{longtable}{|l|l|r|r|r|l|l|l|}
\hline
\multicolumn{8}{|c|}{Financial Crisis} \\
\hline
Country & Segment & Mean & Variance & Skewness & ERM & ERM~II & Euro \\
\hline
<<echo=FALSE, results=tex>>=
writeLines(unlist(lapply(c(11, 19), format_moments)))
@
\hline
\multicolumn{8}{|c|}{The Eastern countries} \\
\hline
<<echo=FALSE, results=tex>>=
writeLines(unlist(lapply(c(3,5,10,15,17,18), format_moments)))
@
\hline
\multicolumn{8}{|c|}{No change countries} \\
\hline
<<echo=FALSE, results=tex>>=
writeLines(unlist(lapply(c(2,4,8,14), format_moments)))
@
\hline
\multicolumn{8}{|c|}{The 90s} \\
\hline
<<echo=FALSE, results=tex>>=
writeLines(unlist(lapply(c(20, 21), format_moments)))
@
\hline
\multicolumn{8}{|c|}{Phase 1 convergence} \\
\hline
<<echo=FALSE, results=tex>>=
writeLines(unlist(lapply(c(1, 6, 7), format_moments)))
@
\hline
\multicolumn{8}{|c|}{Phase 2 convergence} \\
\hline
<<echo=FALSE, results=tex>>=
writeLines(unlist(lapply(c(9,12,13,16), format_moments)))
@
\caption{Break point, parameter estimates and ERM information}
\label{resulttable}
\end{longtable}

@@

\readme{Jesus: include p-values in table; no place left; Jesus: opening seems a bit confusing and obscure} Table \ref{resulttable} presents the break dates and the parameter values of the segments as well as the dates of entry into the ERM/ERM~II and the date of the euro introduction.

@@

The first group that draws particular attention is a subgroup of the PIIGS (Portugal, Ireland, Italy, Greece and Spain) countries consisting of Ireland and Spain. Both show a break in the summer of 2008 following the financial crisis and the severe problems in both countries concerning the real estate situation. In both countries this had a strong deflationary impact, with Ireland actually showing negative inflation rates for an extended period of time; in Spain, which was affected by a housing bubble of roughly the same size as Ireland, the housing prices did not adjust so abruptly and in such a degree. This deflationary trend, which for a time affected all countries, can be seen in Figure \ref{greyplot}, which depicts a rolling mean estimation, where the mean is calculated using the first moment of a fitted GL distribution. There is evidence of a decline in mean inflation, which returns to its previous path roughly a year later. \\
The second group consists of the Eastern European countries: Czech Republic, Estonia, Hungary, Poland, Slovakia and Slovenia. In almost all of these countries, the mean and the variance of their inflation rates declined in the later part of the 90s (Slovenia is an exception with an increase in variance after the break). Most of them experienced a break in 1997/1998, with Poland, Slovakia and Slovenia somewhat later. The break during the late 90s is a result of these countries efforts to curb in inflation by decreasing money supply and indicates the time when these countrie's actually overcame the biggest transitionally shocks on their way towards free market economies. Slovakia is very similar to Slovenia, as its inflation dynamics too changed roughly one year prior to its entry into ERM~II. \\
Another interesting group consists -- with the exception of Austria and Luxembourg -- of the countries of the former DM--zone. All these countries have a very low inflation rate in both mean and variance and their inflation dynamics did not change. \\
The next group consists of two countries that decided not to be part of any European Monetary System\footnote{If we ignore the short participating time of both countries in the EMS.}, the United Kingdom and Sweden. Both show a clear break in the early 90s. They can be traced back to economic crisis, the currency crisis of the United Kingdom cumulating into the ``Black Wednesday'' in 1992 and the run on the Swedish currency in the same year. \\
Austria, Finland and France adjusted to the requirements of the EMU during the first phase of convergence. These are also the countries with the lowest inflation rates in both mean and variance after the adjustment. \\
The last group consists of three PIIGS countries and Luxembourg. Greece and Portugal adjusted to the ERM shortly before or afterwards its introduction. Italy was successfull in reducing its inflation rate, whereas Luxembourg experienced a doubling of its former -- very low -- inflation rate with the variance rising by a factor of 14 with a strong shift in skewness towards the left.

@@

<<echo=FALSE>>=
data("hicps", package = "glogis")
euhi <- hicps[, colnames(hicps) == "EU"]
euhi <- na.omit(euhi)
euhim1 <- mean(window(euhi, start=yearmon(1990), end= yearmon(1998+11/12)))
euhim2 <- mean(window(euhi, start=yearmon(1998), end=yearmon(2008+7/12)))
euhim3 <- mean(window(euhi, start=yearmon(2008+8/10), end=yearmon(2010+11/12)))
euhiv1 <- var(window(euhi, start=yearmon(1990), end= yearmon(1998+11/12)))
euhiv2 <- var(window(euhi, start=yearmon(1998), end=yearmon(2008+7/12)))
euhiv3 <- var(window(euhi, start=yearmon(2008+8/10), end=yearmon(2010+11/12)))
euhis1 <- skewness(window(euhi, start=yearmon(1990), end= yearmon(1998+11/12)))
euhis2 <- skewness(window(euhi, start=yearmon(1998), end=yearmon(2008+7/12)))
euhis3 <- skewness(window(euhi, start=yearmon(2008+8/10), end=yearmon(2010+11/12)))
@

@@

@@

Regarding the general trend, Table \ref{grouptable} presents lower, medium and upper (25, 50, 75) percentiles of the first three moments of the inflation series in January 1994, 1998, 2007 and 2010. The countries are grouped into the ``core'' of the eurozone (Austria, Belgium, Finland, France, Germany, Luxembourg and Netherlands), the PIIGS countries (Portugal, Italy, Ireland, Greece and Spain) and all the others. For this, we took account of possible breaks in the moments.  \\

\begin{longtable}{|l|l|r|r|r|r|}
\hline
\multicolumn{6}{|c|}{Core--euro} \\
\hline
Moment & Quantil & 1994 & 1998 & 2007 & 2010\\
\hline
<<echo=FALSE, results=tex>>=
euro <- c(1, 2, 6, 7, 8, 13, 14)
others <- c(3, 4, 5, 10, 15, 17, 18, 20, 21)
piigs <- c(9, 11, 12, 16, 19)
writeLines(format_cross_country(cross_country_moments(euro, date=yearmon(c(1994,1998,2007,2010)))))
@
\hline
\multicolumn{6}{|c|}{Others} \\
\hline
<<echo=FALSE, results=tex>>=
writeLines(format_cross_country(cross_country_moments(others, date=yearmon(c(1994,1998,2007,2010)))))
@
\hline
\multicolumn{6}{|c|}{PIIGS} \\
\hline
<<echo=FALSE, results=tex>>=
writeLines(format_cross_country(cross_country_moments(piigs, date=yearmon(c(1994,1998,2007,2010)))))
@
\hline
\caption{Development of first three moments}
\label{grouptable}
\end{longtable}

In the core--zone, we see a decrease in the lower quartil from 1994 to 2010, with no change in the medium inflation rate, which is due to the fact, that there is no break after 1998 for these countries; however, the interquartile range decreased by almost 20\%. The variance increased slightly (but with a 15\% decrease in interquartile range), whereas skewness stayed symmetric in the medium, with almost no change in the interquartile range. \\
The other countries were successfull in decreasing all three moments, but they are still higher than the euro--zone countries. Most importantly, the interquartile range for all moments decreased considerably showing a clear convergence towards the core--zone countries.  The PIIGS countries decreased inflation rates as well with a strong change of skewness towards more negative values. The interquartile range of their inflation variance increased considerably, which is mostly due to Spain and Ireland and their strong reaction to the financial crisis.
